\chapter{Introduction}
% Your report should start with an introduction chapter that motivates the subject in general and describes the problem you are trying to solve.

Embedded computing systems are being deployed on an increasingly larger scale, and it is expected that this trend will continue in the forseable future. It follows that there is a great market for developing software for embedded computing systems. In contrast to general-purpose computing, embedded computing often places strong constraints on energy consumption, performance, and cost. In order to implement these qualities in a system, software developers must make extensive use of low-level hardware functionality, and specific knowledge of the hardware components and technology used is required. 

The course TDT4258 Energy Efficient Computer Systems gives an introduction to microcontroller programming on a practical and theoretical level, with a strong focus on energy-efficiency. Through lectures and three comprehensive exercises the students get theoretical knowledge as well as hands-on experience. The development platform used is the EFM32GG-DK3750 development kit from Silicon Labs, and the tools used include energyAware Commander and the GNU Compiler Collection.

In the first exercise we will connect the prototyping board to a specially built gamepad with LEDs and buttons, and write a small program that lets the user to control the gamepad LEDs through the buttons. The purpose of the exercise is to introduce the development kit hardware, to become familiar with writing low-level energy-efficient software, and to learn software development using the GNU-toolchain. All code will be written in assembly using the Thumb-2 instruction set, and no operating system will be running. An important exercise goal is that the program should be energy-efficient, and thus we need to utilize low-level hardware functions to improve the power consumption. To learn different techniques to improve the power consumption, we are encouraged to read the relevant reference manuals and data sheets. 

